---
title: "Homework 5"
author: "PSTAT 115, Winter 2025"
date: "Due on December, 2025 at 11:59 pm"
format: pdf
---

```{r setup, include=TRUE}
cmdstanr::register_knitr_engine(override = FALSE)
options(tinytex.verbose = TRUE)
options(buildtools.check = function(action) TRUE )
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(testthat))
library(coda)


```


\section*{Problem 1.  Logistic regression for toxicity data}

### Logistic regression for pesticide toxicity data.

A environmental agency is testing the effects of a pesticide that can cause acute poisoning in bees, the world's most important pollinator of food crops. The environmental agency collects data on exposure to different levels of the pestidicide in parts per million (ppm).  The agency also identifies collapsed beehives, which they expect could be due to acute pesticide poisoning.  In the data they collect, each observation is pair $(x_i, y_i)$, where $x_i$ represents the dosage of the pollutant and $y_i$ represents whether or not the hive survived.  Take $y_i=1$ means that the beehive has collapsed from poisoning and $y_i=0$ means the beehive survived.  The agency collects data at several different sites, each of which was exposed to a different dosages. The resulting data can be seen below:


```{r echo=FALSE}
inv_logit <- function(x) { exp(x)/(1 + exp(x)) }
x <- round(runif(20, 1, 2), 2)
theta <- inv_logit(-5 + 4*x)
y <- rbinom(length(x), 1, theta)
```


```{r logistic_reg_setup}
x <- c(1.06, 1.41, 1.85, 1.5, 0.46, 1.21, 1.25, 1.09, 
       1.76, 1.75, 1.47, 1.03, 1.1, 1.41, 1.83, 1.17, 
       1.5, 1.64, 1.34, 1.31)
    
y <- c(0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 
       1, 0, 0, 1, 1, 0, 0, 1, 1, 0)
```

Assume that beehive collapse, $y_i$, given pollutant exposure level $x_i$, is $Y_i \sim \text{Bernoulli}(\theta(x_i))$, where $\theta(x_i)$ is the probability of death given dosage $x_i$.  We will assume that $\text{logit}(\theta_i(x_i)) = \alpha + \beta x_i$ where $\text{logit}(\theta)$ is defined as $\text{log}(\theta / (1-\theta))$. This model is known as _logistic regression_ and is one of the most common methods for modeling probabilities of binary events.  

**1a.** Solve for $\theta_i(x_i)$ as a function of $\alpha$ and $\beta$ by inverting the logit function.  If you haven't seen logistic regression before (it is covered in more detail in PSTAT 127 and PSTAT131), it is essentially a generalization of linear regression for binary outcomes. The inverse-logit function maps the linear part, $\alpha + \beta x_i$, which can be any real-valued number into the interval [0, 1] (since we are modeling probabilities of binary outcome, we need the mean outcome to be confined to this range).

_Type your answer here, replacing this text._

**1b** The dose at which there is a 50\% chance of beehvive collapse, $\theta(x_i) = 0.5$, is known as LD50 ("lethal dose 50%"), and is often of interest in toxicology studies.  Solve for LD50 as a function of $\alpha$ and $\beta$.  

_Type your answer here, replacing this text._

**1c**  Implement the logistic regression model in stan by reproducing the stan model described here: [https://mc-stan.org/docs/2_18/stan-users-guide/logistic-probit-regression-section.html](https://mc-stan.org/docs/2_18/stan-users-guide/logistic-probit-regression-section.html).  In this model, we assume the improper prior $p(\alpha, \beta) \propto \text{const}$.  Run the stan model on the beehive data to get Monte Carlo samples. Compute Monte Carlo samples of the LD50 by applying the function derived in the previous part to your $\alpha$ and $\beta$ samples. Report and estimate of the posterior mean of the LD50 by computing the sample average of all Monte Carlo samples of LD50. 

```{r stan_logistic, tags=c()}
# YOUR CODE HERE

logistic_fit <- NULL # YOUR CODE HERE
logistic_samples <- NULL # YOUR CODE HERE

alpha_samples <- logistic_samples$alpha
beta_samples <- logistic_samples$beta

ld50 <- NULL # YOUR CODE HERE
print(ld50)
```
Fill in the compute curve function, which computes the probability of hive collapse for each value of $x$ in `xgrid`.  Then run the code below to make a plot showing both 50\% and 95% confidence band for the probability of a hive collapse as a function of pollutant exposure, Pr($y=1 \mid \alpha, \beta, x)$.  This will plot your predicted hive collapse probabilities for dosages from $x = 0$ to $2$. Verify that you computed the LD50 correctly by identifying the x-value at which the posterior mean crosses 0.5.


```{r dependson="stan_logistic", tags=c()}
xgrid <- seq(0, 2, by=0.1)

## Evaluate probability on the xgrid for one alpha, beta sample
compute_curve <- function(sample) {
  alpha <- sample[1]
  beta <- sample[2]
  
  prob <- NULL # YOUR CODE HERE
  prob
  
}

predictions <- apply(cbind(alpha_samples, beta_samples), 1, compute_curve)

quantiles <- apply(predictions, 1, function(x) quantile(x, c(0.025, 0.25, 0.75, 0.975)))
posterior_mean <- rowMeans(predictions)

tibble(x=xgrid, 
       q025=quantiles[1, ], 
       q25=quantiles[2, ], 
       q75=quantiles[3,], 
       q975=quantiles[4, ], 
       mean=posterior_mean) %>% 
  ggplot() + 
  geom_ribbon(aes(x=xgrid, ymin=q025, ymax=q975), alpha=0.2) + 
  geom_ribbon(aes(x=xgrid, ymin=q25, ymax=q75), alpha=0.5) + 
  geom_line(aes(x=xgrid, y=posterior_mean), size=1) + 
  geom_vline(xintercept = ld50, linetype="dashed") +
  geom_hline(yintercept = 0.5, linetype="dashed") + 
  theme_bw(base_size=16) + ylab("Probability of hive collapse") + xlab("Dosage")
```
### Problem 2. Estimating Skill In Baseball
In baseball, the batting average is defined as the fraction of base hits (successes) divided by "at bats" (attempts).  We can conceptualize a player's "true" batting skill as $p_i = \lim_{n_i\to\infty} \frac{y_i}{n_i}$.  In other words, if each at bat was independent (a simplifying assumption), $p_i$ describes the total fraction of success for player $i$ as the number of attempts gets very large.  Our goal is to estimate the true skill of all player as best as possible using only a limited amount of data.  As usual, for independent counts of success/fail data it is reasonable to assume that $Y_i \sim \text{Bin}(n_i, p_i)$.  The file "lad.csv" includes the number of hits, `y` and the number of attempts `n` for $J=10$ players on the Los Angeles Dodgers after the first month of the 2019 baseball season.  The variable `val` includes the end-of-season batting average and will be used to validate the quality of various estimates. If you are interested, at the end of the assignment we have included the code that was used to scrape the data.  


```{r baseball_setup}

baseball_data <- read_csv("lad.csv", col_types=cols())
baseball_data

## observed hits in the first month
y <- baseball_data$y

## observed at bats in the first month
n <- baseball_data$n

## observed batting average in the first month (same as MLE)
theta_mle <- y/n

## number of players 
J <- nrow(baseball_data)

## end of the year batting average, used to evaluate estimates
val <- baseball_data$val

```


**2a.**  Compute the standard deviation of the empirical batting average, $y/n$ and then compute the sd of the "true skill", (the `val` variable representing the end of season batting average).  Which is smaller? Why does this make sense? _Hint:_ What sources of variation are present in the empirical batting average?


```{r tags=c()}
empirical_sd <- NULL # YOUR CODE HERE
true_sd <- NULL # YOUR CODE HERE
print(empirical_sd)
print(true_sd)
```
_Type your answer here, replacing this text._

```{r}
. = ottr::check("tests/q3a.R")
```

**2b**.  Consider two estimates for the true skill of player $i$, $p_i$: 1) $\hat p_i^{(\text{mle})} = \frac{y_i}{n_i}$ and 2) $\hat p_i^{(\text{comp})} = \frac{\sum_j y_j}{\sum n_j}$.  Estimator 1) is the MLE for each player and ignores any commonalities between the observations. This is sometimes termed the "no pooling" estimator since each parameter is estimating separately without "pooling" information between them.  Estimator 2) assumes all players have identical skill and is sometimes called the "complete pooling" estimator, because the data from each problem is completely "pooled" into one common set. In this problem, we'll treat the end-of-season batting average as a proxy for true skill, $p_i$.  Compute the root mean squared error (RMSE), $\sqrt{\frac{1}{J}\sum_i (\hat p_i - p_i)^2}$ for the "no pooling" and "complete pooling" estimators using the variable `val` as a stand-in for the true $p_i$.  Does "no pooling" or "complete pooling" give you a better estimate of the end-of-year batting averages in this specific case?
    

```{r dependson="baseball_setup", tags=c()}

# Maximum likelihood estimate
phat_mle <- NULL # YOUR CODE HERE

# Pooled estimate
phat_pooled <- NULL # YOUR CODE HERE

rmse_complete_pooling <- NULL # YOUR CODE HERE
rmse_no_pooling <- NULL # YOUR CODE HERE

print(sprintf("MLE: %f", rmse_no_pooling))
print(sprintf("Pooled: %f", rmse_complete_pooling))

```
_Type your answer here, replacing this text._

The no pooling and complete pooling estimators are at opposite ends of a spectrum.  There is a more reasonable compromise: "partial pooling" of information between players.  Although we assume the number of hits follow a binomial distribution.  To complete this specification, we assume $\text{logit}(p_i) \sim N(\mu, \tau^2)$ for each player $i$.  $\mu$ is the "global mean" (on the logit scale), $\text{exp}(\mu)/(1+\text{exp}(\mu)$ is the overall average batting average across all players.  $\tau$ describes how much variability there is in the true skill of players.  If $\tau = 0$ then all players are identical and the only difference in the observed hits is presumed to be due to chance.  If $\tau^2$ is very large then the true skill differences between players is assumed to be large and our estimates will be close to the "no pooling" estimator.  How large should $\tau$ be? We don't know but we can put a prior distribution over the parameter and sample it along with the $p_i$'s!  Assume the following model:

$$\begin{aligned}
& y_i \sim Bin(n_i, p_i) \\
& \theta_i = logit(p_i)\\
&\theta \sim N(\mu, \tau^2)\\
&p(\mu) \propto \text{const} \\
&p(\tau) \propto \text{Cauchy}(0, 1)^+ \text{, (the Half-cauchy distribution, see part d.)}
\end{aligned}$$

```{r}
. = ottr::check("tests/q3b.R")
```

**2c.**  State the correct answer in each case: as $\tau \to \infty$, the posterior mean estimate of $p_i$ in this model will approach the (complete pooling / no pooling) estimator and as $\tau \to 0$ the posterior mean estimate of $p_i$ will approach the  (complete pooling / no pooling) estimator.  Give a brief justification for your answer.    

_Type your answer here, replacing this text._

**2d.**  Implement the hierarchical binomial model in Stan.  As a starting point for your Stan file modify the `eight_schools.stan` file we have provided and save it as `baseball.stan`. To write the hierarchical binomial model, we need the following modifications to the normal hierarchical model:

   - Since we are fitting a hierarchical binomial model, not a normal distribution, we no longer need sampling variance $\sigma_i^2$.  Remove this from the data block. 
    
   - The outcomes `y` are now integers.  Change `y` to an array of integer types in the data block.
    
   - We need to include the number of at bats for each player (this is part of the binomial likelihood).  Add an array of integers, `n` of length $J$ to the data block.
    
   - Replace the sampling model for $y$ with the binomial-logit: `binomial_logit(n, theta)`. This is equivalent to `binomial(n, inv_logit(theta))`.
    
   - The model line for `eta` makes $\theta_i \sim N(\mu, \tau^2)$.  Leave this in the model.
    
   - Add a half-cauchy prior distribution for $\tau$: `tau ~ cauchy(0, 1);`. The half-cauchy has been suggested as a good default prior distribution for group-level standard deviations in hierarchical models.  See \url{http://www.stat.columbia.edu/~gelman/research/published/taumain.pdf}.

Find the posterior means for each of the players batting averages by looking at the samples for `inv_logit(theta_samples)`. Report the RMSE for hierarchical estimator.  How does this compare to the RMSE of the complete pooling and no pooling estimators?  Which estimator had the lowest error? 

```{r baseball_stan, dependson="baseball_setup", tags=c()}

# Run Stan and compute the posterior mean

stan_data <- list(J=J, n=n, y=y)
baseball_model <- cmdstan_model("baseball.stan")
baseball_results <- baseball_model$sample(data=stan_data, iter_sampling=2000, refresh=0)
baseball_draws <- baseball_results$draws(format="df")

# Get the matrix of Theta samples for all players
# Should be a matrix of size num_draws * num_players
# where num_players is 10
# Note: theats are on logit scale, need to convert back to prob
theta_samples <- NULL # YOUR CODE HERE

# Get batting averages by inverting with this function
inv_logit <- function(x) {
  # YOUR CODE HERE
}
# and compute the posterior mean for each theta
pm <- NULL # YOUR CODE HERE


# RMSE From Stan posterior means
rmse_partial_pooling <- NULL # YOUR CODE HERE

print(c(rmse_complete_pooling, rmse_no_pooling, rmse_partial_pooling))
```
```{r}
. = ottr::check("tests/q3d.R")
```

**2e.**  Use the `shrinkage_plot` function provided below to show how the posterior means shrink the empirical batting averages.  Pass in `y/n` and the posterior means of $p_i$ as arguments.


```{r echo=TRUE, tags=c()}
shrinkage_plot <- function(empirical, posterior_mean,
                           shrink_point=mean(posterior_mean)) {
  
  tibble(y=empirical, pm=posterior_mean) %>% 
    ggplot() + 
    geom_segment(aes(x=y, xend=pm, y=1, yend=0), linetype="dashed") + 
    geom_point(aes(x=y, y=1)) + 
    geom_point(aes(x=pm, y=0)) + 
    theme_bw(base_size=16) + 
    geom_vline(xintercept=shrink_point, color="blue", size=1.2) + 
    ylab("") + xlab("Estimate") + 

    scale_y_continuous(breaks=c(0, 1), 
                       labels=c("Posterior Mean", "MLE"), 
                       limits=c(0,1))

}

# YOUR CODE HERE
```


```{r dependson="baseball_stan", tags=c()}
# YOUR CODE HERE
```

_Type your answer here, replacing this text._

### Appendix: Code for scraping Dodgers baseball data

[http://billpetti.github.io/baseballr/](http://billpetti.github.io/baseballr/)

```{r echo=TRUE, eval=FALSE}

## Install the baseballr package
devtools::install_github("BillPetti/baseballr")

library(baseballr)
library(tidyverse)

## Download data from the chosen year
year <- 2025

one_month <- daily_batter_bref(t1 = sprintf("%i-04-01", year), t2 = sprintf("%i-05-01", year))
one_year <- daily_batter_bref(t1 = sprintf("%i-04-01", year), t2 = sprintf("%i-10-01", year))

## filter to only include players who hat at least 10 at bats in the first month
one_month <- one_month %>% filter(AB > 10)
one_year <- one_year %>% filter(Name %in% one_month$Name)

one_month <- one_month %>% arrange(Name)
one_year <- one_year %>% arrange(Name)

## Look at only the Dodgers
LAD <- one_year %>% filter(Team == "Los Angeles" & Level == "MLB-NL") %>% .$Name

lad_month <- one_month %>% filter(Name %in% LAD)
lad_year <- one_year %>% filter(Name %in% LAD)

write_csv(tibble(name=lad_month$Name,
                 y=lad_month$H,
                 n=lad_month$AB,
                 val=lad_year$BA), 
          path="lad.csv")

```
